Low-cost linear-scaling \emph{ab initio} molecular dynamics for weakly-interacting systems

### 2017-09-15 almo-langevin.tex
Added section: Supplemental Materials and moved ALMO SCF delta force comparison to there. Added linear scale timing plot and mean-squared displacement plot to SI.

To adapt ALMO DFT for AIMD, the forces on atoms were computed \emph{approximately} using the idempotent DM obtained from ALMOs in a straightforward procedure that computes only the Hellmann-Feynman and Pulay forces while ignoring non-SCF forces (see the Supplemental Material). 
The difference between these forces and the exact forces obtained using perfectly-converged fully-delocalized KS orbitals is denoted $\delta f_{i\alpha}(t)$ 
%
\begin{align}
\label{eq:deltaf}
f^{\text{KS}}_{i\alpha}(t) = f^{\text{ALMO}}_{i\alpha}(t) + \delta f_{i\alpha} (t),
\end{align}
%
where $\alpha$ is a Cartesian component of the force acting on atom $i$ at time $t$. $\delta f_{i\alpha} (t)$ comprises neglected terms that originate from %fixed terms in the trial ALMOs (i.e. orbitals obtained in the first SCF step are not re-optimized in the second step), 
a finite localization radius $R_c$ and incomplete SCF optimization of the electronic degrees of freedom. 
$\delta f_{i\alpha} (t)$ can be reduced \emph{systematically} to zero by increasing $R_c$ (Figure~\ref{fig:forcecomp}) and decreasing $\epsilon_{\text{SCF}}$ -- the maximum norm of the gradient of the \mbox{ALMO} DFT energy with respect to the ALMO coefficients. $\epsilon_{\text{SCF}}$ is chosen such that the electronic ground state is considered sufficiently close to the Born-Oppenheimer state for values less than $\epsilon_{\text{SCF}}$.
%ZZZ For $\epsilon_{\text{SCF}} \leqslant 10^{-7}$~a.u. ALMO SCF is considered converged.

\begin{figure}[t]
\includegraphics[trim={0cm 0cm 0.1cm 0.1cm},clip,width=8.6cm]{DeltaForceComparison_ALMO_SCF.eps}
\caption{\label{fig:forcecomp} Dependence of the Euclidean norm of the instantaneous error $\langle \| \delta \vec{f_{i}}(t) \| \rangle_{i}$ (red line) on the ALMO localization radius $R_c$ expressed in units of atomic van der Waals radii. Black line is the magnitude of the time average of the instantaneous error vector. %calculated as the average difference between fully converged Born-Oppenheimer forces and fully converged ALMO DFT forces, averaged over all dimensions and all molecules for each time step. 
Forces are computed with the PBE XC functional, TZV2P basis set, and $\epsilon_{\text{SCF}} = 10^{-7}$~a.u. for the configurations from a 10-ps trajectory generated with KS-based AIMD at T=298~K.}
\end{figure}

@@

In order to confirm that ALMO DFT forces are accurate in the limit where localization radius $R_{c} \rightarrow \infty$, we computed the average difference in force between fully converged Born-Oppenheimer forces and forces calculated from ALMO DFT with varying $R_{c}$. 
Figure \ref{fig:forcecomp} shows that as $R_{c}$ is increased to infinity, forces converge toward the Born-Oppenheimer limit.

To test the accuracy of our unconverged ALMO based MD we computed the Infrared (IR) spectrum, radial distribution function (RDF), kinetic energy distribution, and mean-squared displacement (MSD) function from a 10 ps MD trajectory (see fig. \ref{fig:dynproperties} and the Supplemental Material). 
We compared the IR spectrum, RDF, and MSD function to corresponding reference functions calculated with fully converged KS DFT forces.
The kinetic energy distribution is compared with the theoretical Maxwell-Boltzmann distribution at the average temperature of the system.
From the slope of the MSD functions we also calculated diffusion constants.
Using fully converged KS DFT forces we calculated D = $\SI{5.47e-9}{\m^{2}\per\s}$, whereas with unconverged ALMO forces we calculated D = $\SI{5.55E-09}{\m^{2}\per\s}$.

@@

%\begin{figure}
%\includegraphics[trim={1.7cm 0.5cm 3.3cm 0.1cm},clip,width=8.6cm]{strongscaling_linear.eps}
%\caption{\label{fig:strongscaling_linear} Linear plots showing LS behaviour of all ALMO methods for liquid water at 298 K. 
%Calculations were performed with the PBE $\mathrm{E_{XC}}$ functional and TZV2P basis set on 256 cores. 
%The localization radius was set to $R_{c} = 1.6$ vdWR.}
%\end{figure}

@@

%\textbf{Supplemental Material}

@@

>>>>

To adapt ALMO DFT for AIMD, the forces on atoms were computed \emph{approximately} using the idempotent DM obtained from ALMOs in a straightforward procedure that computes only the Hellmann-Feynman and Pulay forces while ignoring non-SCF forces (see the Supplemental Material). 
The difference between these forces and the exact forces obtained using perfectly-converged fully-delocalized KS orbitals is denoted $\delta f_{i\alpha}(t)$ 
%
\begin{align}
\label{eq:deltaf}
f^{\text{KS}}_{i\alpha}(t) = f^{\text{ALMO}}_{i\alpha}(t) + \delta f_{i\alpha} (t),
\end{align}
%
where $\alpha$ is a Cartesian component of the force acting on atom $i$ at time $t$. $\delta f_{i\alpha} (t)$ comprises neglected terms that originate from %fixed terms in the trial ALMOs (i.e. orbitals obtained in the first SCF step are not re-optimized in the second step), 
a finite localization radius $R_c$ and incomplete SCF optimization of the electronic degrees of freedom. 
$\delta f_{i\alpha} (t)$ can be reduced \emph{systematically} to zero by increasing $R_c$ (Figure~\ref{fig:forcecomp} in the Supplemental Material) and decreasing $\epsilon_{\text{SCF}}$ -- the maximum norm of the gradient of the \mbox{ALMO} DFT energy with respect to the ALMO coefficients. $\epsilon_{\text{SCF}}$ is chosen such that the electronic ground state is considered sufficiently close to the Born-Oppenheimer state for values less than $\epsilon_{\text{SCF}}$.
%ZZZ For $\epsilon_{\text{SCF}} \leqslant 10^{-7}$~a.u. ALMO SCF is considered converged.

@@

In order to confirm that ALMO DFT forces are accurate in the limit where localization radius $R_{c} \rightarrow \infty$, we computed the average difference in force between fully converged Born-Oppenheimer forces and forces calculated from ALMO DFT with varying $R_{c}$. 
Figure \ref{fig:forcecomp} in the Supplemental Material shows that as $R_{c}$ is increased to infinity, forces converge toward the Born-Oppenheimer limit.

To test the accuracy of our unconverged ALMO based MD we computed the Infrared (IR) spectrum, radial distribution function (RDF), kinetic energy distribution, and mean-squared displacement (MSD) function from a 10 ps MD trajectory (see fig. \ref{fig:dynproperties} and \ref{fig:diffusion} in the Supplemental Material). 
We compared the IR spectrum, RDF, and MSD function to corresponding reference functions calculated with fully converged KS DFT forces.
The kinetic energy distribution is compared with the theoretical Maxwell-Boltzmann distribution at the average temperature of the system.
From the slope of the MSD functions we also calculated diffusion constants.
Using fully converged KS DFT forces we calculated D = $\SI{5.47e-9}{\m^{2}\per\s}$, whereas with unconverged ALMO forces we calculated D = $\SI{5.55E-09}{\m^{2}\per\s}$.

@@

@@

@@

\clearpage
\widetext
\setcounter{figure}{0}
\setcounter{page}{1}
\renewcommand{\thefigure}{S\arabic{figure}}
\begin{center}
\textbf{\large Supplemental Material}
\end{center}

\begin{figure}[!htb]
\includegraphics[trim={0cm 0cm 0.1cm 0.1cm},clip,width=\textwidth]{DeltaForceComparison_ALMO_SCF.eps}
\caption{\label{fig:forcecomp} Dependence of the Euclidean norm of the instantaneous error $\langle \| \delta \vec{f_{i}}(t) \| \rangle_{i}$ (red line) on the ALMO localization radius $R_c$ expressed in units of atomic van der Waals radii. Black line is the magnitude of the time average of the instantaneous error vector. %calculated as the average difference between fully converged Born-Oppenheimer forces and fully converged ALMO DFT forces, averaged over all dimensions and all molecules for each time step. 
Forces are computed with the PBE XC functional, TZV2P basis set, and $\epsilon_{\text{SCF}} = 10^{-7}$~a.u. for the configurations from a 10-ps trajectory generated with KS-based AIMD at T=298~K.}
\end{figure}

\begin{figure}[!htb]
\includegraphics[trim={1.7cm 0.5cm 3.3cm 0.1cm},clip,width=\textwidth]{strongscaling_linear.eps}
\caption{\label{fig:strongscaling_linear} Linear plots showing LS behaviour of all ALMO methods for liquid water at 298 K. 
Calculations were performed with the PBE $\mathrm{E_{XC}}$ functional and TZV2P basis set, averaged over 10 MD steps. 
The localization radius was set to $R_{c} = 1.6$ vdWR and $\epsilon_{\text{SCF}} = 10^{-2}$~a.u.}
\end{figure}

\begin{figure}[!htb]
\includegraphics[trim={2.5cm 0.2cm 4.5cm 1.2cm},clip,width=\textwidth]{Diffusion_Comparison.eps}
\caption{\label{fig:diffusion} Calculated mean-squared displacement function of water using ALMO DFT with $\epsilon_{\text{SCF}} = 10^{-2}$~a.u. and $R_{c} = 1.6$ vdWR.
Compared with equivalent MD run using OT DFT as a reference.
Calculations were performed at 298 K with the PBE $\mathrm{E_{XC}}$ functional and TZV2P basis set, using 125 water molecules for 10 ps.}
\end{figure}
